\chapter{Treatment System}
\label{ch:treatment}

\begin{tcolorbox}[colback=treatment!10, colframe=treatment!50, title=\textbf{Digital Therapeutics}]
\textit{How do you model a drug that doesn't exist yet? How do you test combination therapies that would take decades to validate clinically? This chapter reveals how our treatment system transforms pharmaceutical mechanisms into computational interventions, enabling virtual clinical trials that can explore therapeutic landscapes impossible to navigate in the real world.}
\end{tcolorbox}

The modern treatment of kidney cancer represents one of medicine's great success stories—the transformation of a universally fatal disease into one where many patients achieve long-term survival. This revolution came through understanding two key principles: unleashing the immune system's natural cancer-fighting ability, and cutting off the tumor's blood supply.

Our treatment system captures both strategies through detailed modeling of immune checkpoint inhibitors (ICIs) and tyrosine kinase inhibitors (TKIs), along with their synergistic combination. But this isn't just about implementing known drugs—it's about creating a platform for exploring future therapeutic possibilities.

\begin{figure}[H]
\centering
\input{figures/treatment_mechanisms}
\caption{Comprehensive treatment mechanisms across ICI, TKI, and combination therapy. \textbf{Left panel:} ICI therapy blocks the PD-1/PD-L1 checkpoint interaction, preventing tumor-mediated immune suppression and reactivating exhausted T cells. \textbf{Middle panel:} TKI therapy targets VEGF-mediated angiogenesis, disrupting tumor blood supply while releasing neoantigens for immune presentation. \textbf{Right panel:} Combination therapy achieves synergy—TKI creates optimal conditions for immune infiltration while releasing tumor antigens, then ICI removes the brakes on immune responses, resulting in enhanced anti-tumor activity that exceeds either monotherapy alone.}
\label{fig:treatment_mechanisms}
\end{figure>

%% ====================================================================
\section{Drug Architecture}
\label{sec:drug-architecture}

The treatment system follows a composition-based design in which individual drugs are combined into a treatment regimen.

\subsection{Abstract Drug Base Class}

All drugs inherit from the abstract \texttt{Drug} base class (located in \texttt{src/treatments/drug.py}), which defines a single interface method:

\begin{lstlisting}[caption={Abstract Drug base class.},label={lst:drug-base}]
class Drug(ABC):
    def __init__(self, model):
        self.model = model

    def step(self, proportion=1.0):
        raise NotImplementedError
\end{lstlisting}

\noindent The \texttt{proportion} parameter controls what fraction of the treatment's total capacity is allocated to this drug.  Each drug receives a reference to the model, granting access to all agent populations and parameters.

\subsection{Treatment Composition}

The \texttt{Treatment} class (in \texttt{src/treatments/treatment.py}) holds a list of \texttt{Drug} instances and distributes the treatment capacity equally among them:

\begin{lstlisting}[caption={Treatment composition with proportional averaging.},label={lst:treatment}]
class Treatment:
    def __init__(self, drugs):
        self.drugs = drugs

    def step(self):
        if not self.drugs:
            return
        for drug in self.drugs:
            drug.step(proportion=1.0 / len(self.drugs))
\end{lstlisting}

\noindent With a single drug, \texttt{proportion} $= 1.0$; with two drugs (e.g., ICI + TKI), each receives \texttt{proportion} $= 0.5$.  This proportional averaging models the clinical reality that combination therapy involves dose adjustments.

\subsection{Treatment Activation}

Treatment does not begin at the start of the simulation.  Instead, it is activated after a configurable number of steps specified by the \texttt{treatment\_start} patient parameter (default: step 100).  Before this time point, the model simulates unperturbed tumor--immune dynamics, establishing a baseline disease state.  At each step from \texttt{treatment\_start} onward, \texttt{Treatment.step()} is called:

\begin{equation}
\label{eq:treatment-activation}
\text{Treatment active at step } t \iff t \geq t_{\text{start}}.
\end{equation}

\noindent The available treatment configurations, selected via the \texttt{treatment} patient parameter, are:

\begin{itemize}[nosep]
  \item \texttt{"None"} -- no drugs (control),
  \item \texttt{"ICI"} -- immune checkpoint inhibitor monotherapy,
  \item \texttt{"TKI"} -- tyrosine kinase inhibitor monotherapy,
  \item \texttt{"ICI+TKI"} -- combination therapy.
\end{itemize}

%% ====================================================================
\section{Immune Checkpoint Inhibitors (ICI)}
\label{sec:ici}

\subsection{Biological Basis}

The programmed death-1 (PD-1) receptor is expressed on activated T cells and serves as a negative regulator of T cell function.  Its ligand, PD-L1 (encoded by the CD274 gene), is frequently upregulated on tumor cells in renal cell carcinoma.  When PD-L1 on a tumor cell engages PD-1 on a T cell, the T cell's cytotoxic activity is suppressed---a process known as immune checkpoint inhibition~\cite{pardoll2012}.

Anti-PD-1 and anti-PD-L1 monoclonal antibodies (collectively termed ICI) block this interaction, restoring T cell effector function.  In clinical practice, ICI therapy (e.g., nivolumab, pembrolizumab) has shown significant survival benefits in RCC patients~\cite{rini2019,motzer2019}.

\subsection{Implementation}

The \texttt{ICIDrug} class (in \texttt{src/treatments/ici.py}) implements three effects per treatment step:

\paragraph{1.\ PD-L1 Blockade on Tumor Cells.}
For each tumor cell, with probability equal to \texttt{proportion} $\times$ \texttt{w\_ici\_effectiveness}:

\begin{equation}
\label{eq:ici-tumor}
P(\text{ICI\_effect} = \texttt{True}) = p \cdot w_{\text{ICI}},
\end{equation}

\noindent where $p$ is the drug proportion and $w_{\text{ICI}}$ is the ICI effectiveness weight (default $1.0$).  When \texttt{ICI\_effect} is \texttt{True}, the tumor cell's \texttt{get\_PD1\_inhibition()} method returns \texttt{False}, effectively disabling the checkpoint pathway for that cell.

\paragraph{2.\ T Cell Activation Restoration.}
For each T cell (across all six T cell subtypes: CD8 cytotoxic, CD8 naive, CD4 naive, CD4 Th1, CD4 Th2, and Treg), with the same probability:

\begin{equation}
\label{eq:ici-tcell}
\texttt{t\_activation\_effect} \leftarrow 1.0.
\end{equation}

\noindent This resets the T cell's activation state to full capacity, counteracting the progressive exhaustion that occurs when T cells encounter PD-L1$^+$ tumor cells without treatment.

\paragraph{3.\ Immune Infiltration Enhancement.}
For immune cell types Th1, Th2, and MastCell, with the same probability, the \texttt{immune\_infiltration\_factor} is incremented:

\begin{equation}
\label{eq:ici-infiltration}
\texttt{immune\_infiltration\_factor} \leftarrow \texttt{immune\_infiltration\_factor} + 1.
\end{equation}

\noindent This models the clinical observation that ICI therapy promotes immune cell infiltration into the tumor mass, as the restored immune microenvironment generates stronger chemotactic signals.

%% ====================================================================
\section{Tyrosine Kinase Inhibitors (TKI)}
\label{sec:tki}

\subsection{Biological Basis}

Vascular endothelial growth factor (VEGF) is a key driver of angiogenesis---the formation of new blood vessels that supply the tumor with nutrients and oxygen.  Renal cell carcinoma is particularly dependent on VEGF signaling due to frequent VHL gene inactivation, making anti-angiogenic therapy a cornerstone of RCC treatment.

Tyrosine kinase inhibitors (e.g., axitinib, sunitinib) block the intracellular signaling domains of VEGF receptors and other receptor tyrosine kinases, thereby inhibiting tumor-driven angiogenesis and directly reducing tumor cell proliferation~\cite{rini2019}.

\subsection{Implementation}

The \texttt{TKIDrug} class (in \texttt{src/treatments/tki.py}) implements four effects:

\paragraph{1.\ Tumor Growth Suppression.}
For each tumor cell, with probability \texttt{proportion} $\times$ \texttt{w\_tki\_effectiveness}, the tumor growth effect is reduced:

\begin{equation}
\label{eq:tki-growth}
\texttt{tumour\_growth\_effect} \leftarrow \texttt{tumour\_growth\_effect} \times (1 - \varepsilon),
\end{equation}

\noindent where $\varepsilon = p \cdot w_{\text{TKI}}$ is the effective TKI strength.

\paragraph{2.\ Angiogenesis Inhibition.}
With the same per-cell probability, the angiogenesis effect is reduced:

\begin{equation}
\label{eq:tki-angio}
\texttt{angiogenesis\_effect} \leftarrow \texttt{angiogenesis\_effect} \times (1 - \varepsilon).
\end{equation}

\noindent This directly suppresses the tumor cell's ability to recruit new blood vessels, reducing nutrient supply and constraining spatial expansion.

\paragraph{3.\ Neoantigen Release.}
With probability $\varepsilon$, the tumor cell's \texttt{TKI\_effect} flag is set to \texttt{True}.  When a tumor cell with \texttt{TKI\_effect = True} subsequently undergoes apoptosis, it releases its neoantigens to nearby dendritic cells (within radius 3), facilitating antigen presentation and downstream T cell activation.  This models the immunogenic cell death induced by TKI therapy.

\paragraph{4.\ Treg Differentiation Suppression.}
For each CD4 naive T cell, with probability $\varepsilon$:

\begin{equation}
\label{eq:tki-treg}
\texttt{treg\_differentiation\_effect} \leftarrow \texttt{treg\_differentiation\_effect} \times (1 - \varepsilon).
\end{equation}

\noindent TKI therapy reduces the differentiation of naive CD4$^+$ T cells into immunosuppressive regulatory T cells (Tregs), thereby shifting the immune balance towards anti-tumor activity.

%% ====================================================================
\section{Combination Therapy}
\label{sec:combination}

\subsection{ICI + TKI}

When both ICI and TKI are administered simultaneously, the \texttt{Treatment} class creates both drug instances, each receiving \texttt{proportion} $= 0.5$:

\begin{equation}
\label{eq:combination}
p_{\text{ICI}} = p_{\text{TKI}} = \frac{1}{2}.
\end{equation}

\noindent The effective strength of each drug is halved compared to monotherapy, reflecting the dose adjustments common in combination regimens.

\subsection{Synergistic Mechanisms}

The combination of ICI and TKI exploits complementary mechanisms that produce synergistic anti-tumor effects:

\begin{enumerate}[nosep]
  \item \textbf{Vascular normalization:} TKI reduces aberrant tumor vasculature, which paradoxically improves blood flow to the tumor center.  This normalized vasculature allows more immune cells to infiltrate the tumor mass.
  \item \textbf{Restored immune function:} ICI blocks the PD-1/PD-L1 checkpoint, restoring the cytotoxic capacity of the infiltrating T cells.  Without ICI, even well-infiltrated T cells may be rendered anergic by PD-L1 engagement.
  \item \textbf{Immunogenic cell death:} TKI-induced tumor cell death releases neoantigens, priming new adaptive immune responses.  ICI ensures that the T cells activated by these neoantigens remain functional.
  \item \textbf{Treg reduction:} TKI suppresses Treg differentiation, reducing the immunosuppressive component of the tumor microenvironment, while ICI restores the activity of effector T cells.
\end{enumerate}

\subsection{Clinical Evidence}

The clinical rationale for ICI + TKI combination therapy in RCC is well established.  The KEYNOTE-426 trial demonstrated that pembrolizumab (anti-PD-1) combined with axitinib (TKI) significantly improved overall survival and progression-free survival compared to sunitinib monotherapy in advanced RCC~\cite{rini2019}.  Similarly, the JAVELIN Renal 101 trial showed that avelumab (anti-PD-L1) plus axitinib improved progression-free survival over sunitinib~\cite{motzer2019}.  The ARON-1 study further examined sex-specific response patterns in this combination setting~\cite{santoni2024}.

The model's treatment system is designed to reproduce the qualitative dynamics observed in these trials: combination therapy produces stronger tumor regression than either monotherapy, and the relative benefit depends on the patient's immune profile and sex-specific parameters (see Chapter~\ref{ch:sex-hormones}).
